---
title: <center><h1>Student Data Modeling for Top 5 and Bottom 5 Countries</h1></center>
author: 
- name: "Nirmal Ghimire, Ph.D."
  url: https://www.linkedin.com/in/nirmal-ghimire-5b96a034/
  affiliation: Watson College of Education, University of North Carolina Wilmington
  affiliation_url: https://www.uncw.edu/ed/
  orcid_id: 0000-0002-2032-1624
date: "`r Sys.Date()`"
output: distill::distill_article
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE,
                      comment = NA,
                      warning = FALSE,
                      message = FALSE,
                      tidy = 'styler',
                      error = FALSE, 
                      highlight = TRUE, 
                     prompt = FALSE)
library(tidyverse)
library(haven)
library(dplyr)
library(data.table)
library(ggplot2)
library(reshape2)
library(naniar)
library(tidyr)
library(psych)
```

## A. Data Structure and Diemnsions

```{r load_data, echo=FALSE}
#Loading required data sets
# Load the saved student_ten data
load("student_ten.RData")

# Subset the data to include only the specified variables
student_ten_subset <- student_ten %>%
  select(CNTRYID, CNT, CNTSCHID, CNTSTUID, OECD, ICTHOME, ICTSCH, GRADE, 
         PAREDINT, WEALTH, ST004D01T, reading_score, math_score, science_score)

# Check the dimensions of the loaded data
dim(student_ten)
dim(student_ten_subset)

# Summary of the subsetted dataset
#summary(student_ten_subset)
#str(student_ten_subset)
# Display the first few rows of the dataset
# names(student_ten)
```


```{r data_scaling, echo=FALSE}
## Data Scaling and Normalization
# 1. Change all variable names to lower case
setnames(student_ten_subset, tolower(names(student_ten_subset)))

# 2. Change the variable class based on attributes

# Convert CNTRYID and CNT to factors
student_ten_subset$cntryid <- as.factor(student_ten_subset$cntryid)
student_ten_subset$cnt <- as.factor(student_ten_subset$cnt)

# Convert CNTSCHID and CNTSTUID to numeric
student_ten_subset$cntschid <- as.numeric(student_ten_subset$cntschid)
student_ten_subset$cntstuid <- as.numeric(student_ten_subset$cntstuid)

# Convert OECD to factor using its labels
student_ten_subset$oecd <- factor(student_ten_subset$oecd, levels = c(0, 1), labels = c("No", "Yes"))

# Convert ICTHOME, ICTSCH to numeric (removing labels)
student_ten_subset$icthome <- as.numeric(student_ten_subset$icthome)
student_ten_subset$ictsch <- as.numeric(student_ten_subset$ictsch)

# Convert GRADE to numeric (removing labels)
student_ten_subset$grade <- as.numeric(student_ten_subset$grade)

# Convert PAREDINT to numeric
student_ten_subset$paredint <- as.numeric(student_ten_subset$paredint)

# Convert WEALTH to numeric
student_ten_subset$wealth <- as.numeric(student_ten_subset$wealth)

# 3. Rename ST004D01T to student_gender and recode it
student_ten_subset <- student_ten_subset %>%
  mutate(student_gender = recode_factor(
    as.factor(st004d01t),
    `1` = "Female",
    `2` = "Male",
    `5` = "Valid Skip",
    `7` = "Not Applicable",
    `8` = "Invalid",
    `9` = "No Response"
  ))

# Drop the old ST004D01T column
student_ten_subset <- student_ten_subset %>% select(-st004d01t)

# Check the structure of the modified data
str(student_ten_subset)
```

## B. Variable Descriptions
* **cntryid:** A categorical variable representing the country ID, where each unique value corresponds to a different country in the dataset.
* **cnt:** A categorical variable representing the country code as a short identifier, such as "USA" for the United States.
* **cntschid:** A numeric variable representing the unique international school ID, which identifies each school within a country.
* **cntstuid:** A numeric variable representing the unique international student ID, which identifies each student within the dataset.
* **oecd:** A categorical variable indicating whether the country is a member of the Organisation for Economic Co-operation and Development (OECD). It has two levels: "Yes" (OECD member) and "No" (non-OECD member).
* **icthome:** A numeric variable representing the availability of Information and Communication Technology (ICT) at the student's home. Higher values indicate greater availability.
* **ictsch:** A numeric variable representing the availability of Information and Communication Technology (ICT) at the student's school. Higher values indicate greater availability.
* **grade:** A numeric variable indicating the student's grade level compared to the modal grade in their country. Negative values indicate below-modal grades, zero indicates the modal grade, and positive values indicate above-modal grades.
* **paredint:** A numeric variable representing the highest level of parental education, measured in international years of schooling.
* **wealth:** A numeric variable representing the family's wealth index, measured using a weighted likelihood estimate (WLE). Higher values indicate greater wealth.
* **student_gender:** A categorical variable representing the student's gender, coded as "Female" or "Male". It also includes categories for cases like "Valid Skip", "Not Applicable", "Invalid", and "No Response".
* **reading_score:** A numeric variable representing the student's weighted average score in reading, calculated from multiple plausible values.
* **math_score:** A numeric variable representing the student's weighted average score in mathematics, calculated from multiple plausible values.
* **science_score:** A numeric variable representing the student's weighted average score in science, calculated from multiple plausible values.

## C. Missing Value Analysis
```{r missing_values, echo=FALSE}
# Check for missing values in the dataset
missing_values <- student_ten_subset %>%
  summarise_all(~sum(is.na(.))) %>%
  gather(variable, missing_count)

# Display the variables with missing values and corresponding percentages
missing_values <- missing_values %>%
  mutate(missing_percentage = missing_count / nrow(student_ten_subset) * 100) %>%
  arrange(desc(missing_percentage))

#missing_values

# Melt the data for visualization
melted_data <- melt(is.na(student_ten_subset))
colnames(melted_data) <- c("Row", "Variable", "Missing")

# Plot missing data
missing_gg <- ggplot(data = melted_data, aes(x = Variable, fill = Missing)) +
  geom_bar() +
  coord_flip() +
  labs(title = "Missing Data Pattern",
       x = "Variables",
       y = "Count of Missing Values") +
  theme_minimal()

# Using naniar package for missing data visualization
gg_miss_var(student_ten_subset) +
  labs(title = "Missing Data Pattern by Variable") +
  theme_minimal()

# Filter out countries with no data
student_ten_filtered <- student_ten_subset %>%
  group_by(cnt) %>%
  filter(sum(!is.na(reading_score) | !is.na(math_score) | !is.na(science_score)) > 0) %>%
  ungroup()

# Dropping unused factor levels
student_ten_filtered$cntryid <- droplevels(student_ten_filtered$cntryid)
student_ten_filtered$cnt <- droplevels(student_ten_filtered$cnt)
```

The missing value analysis shows that the `ictsch` (ICT availability at school) and `icthome` (ICT availability at home) variables have substantial missing data, with 22.83% and 21.77% of values missing, respectively, which could significantly impact the reliability of analyses involving these variables. The `paredint` (highest parental education) variable has 4.17% missing data, while the `wealth` (family wealth index) variable has 2.44% missing, both of which are less concerning but still noteworthy. 

## D. Descriptive Analyiss of Key Variables
### i. Summary Statistics for Students' Reading Scores, Math Scores, Science Scores, Parental Education, and Wealth
```{r summary_statistics, echo=FALSE}
# Calculate extended summary statistics for each numeric variable by country
reading_summary_stats <- student_ten_filtered %>%
   summarise(
    mean_reading = mean(reading_score, na.rm = TRUE),
    median_reading = median(reading_score, na.rm = TRUE),
    sd_reading = sd(reading_score, na.rm = TRUE),
    min_reading = min(reading_score, na.rm = TRUE),
    max_reading = max(reading_score, na.rm = TRUE),
    skewness_reading = skew(reading_score, na.rm = TRUE),
    kurtosis_reading = kurtosi(reading_score, na.rm = TRUE),

    mean_math = mean(math_score, na.rm = TRUE),
    median_math = median(math_score, na.rm = TRUE),
    sd_math = sd(math_score, na.rm = TRUE),
    min_math = min(math_score, na.rm = TRUE),
    max_math = max(math_score, na.rm = TRUE),
    skewness_math = skew(math_score, na.rm = TRUE),
    kurtosis_math = kurtosi(math_score, na.rm = TRUE),

    mean_science = mean(science_score, na.rm = TRUE),
    median_science = median(science_score, na.rm = TRUE),
    sd_science = sd(science_score, na.rm = TRUE),
    min_science = min(science_score, na.rm = TRUE),
    max_science = max(science_score, na.rm = TRUE),
    skewness_science = skew(science_score, na.rm = TRUE),
    kurtosis_science = kurtosi(science_score, na.rm = TRUE),

    mean_paredint = mean(paredint, na.rm = TRUE),
    median_paredint = median(paredint, na.rm = TRUE),
    sd_paredint = sd(paredint, na.rm = TRUE),
    min_paredint = min(paredint, na.rm = TRUE),
    max_paredint = max(paredint, na.rm = TRUE),
    skewness_paredint = skew(paredint, na.rm = TRUE),
    kurtosis_paredint = kurtosi(paredint, na.rm = TRUE),

    mean_wealth = mean(wealth, na.rm = TRUE),
    median_wealth = median(wealth, na.rm = TRUE),
    sd_wealth = sd(wealth, na.rm = TRUE),
    min_wealth = min(wealth, na.rm = TRUE),
    max_wealth = max(wealth, na.rm = TRUE),
    skewness_wealth = skew(wealth, na.rm = TRUE),
    kurtosis_wealth = kurtosi(wealth, na.rm = TRUE)
  ) %>%
  ungroup()

# Pivot the table to make it longer with variables as rows and statistics as columns
reading_summary_stats_long <- reading_summary_stats %>%
  pivot_longer(
    cols = everything(),  # Pivot all columns
    names_to = "Variable_Statistic",  # Combine the variable and statistic names into one column
    values_to = "Value"  # The values go into the "Value" column
  ) %>%
  separate(Variable_Statistic, into = c("Statistic", "Variable"), sep = "_", extra = "merge") %>%
  pivot_wider(
    names_from = Statistic,  # Spread statistics (e.g., mean, median) across columns
    values_from = Value  # Values from the pivoted column
  )

# Print the reshaped table
print(reading_summary_stats_long)


# Calculate summary statistics for each numeric variable by country
country_summary_stats <- student_ten_filtered %>%
  group_by(cnt) %>%
  summarise(
    mean_reading = mean(reading_score, na.rm = TRUE),
    median_reading = median(reading_score, na.rm = TRUE),
    sd_reading = sd(reading_score, na.rm = TRUE),
    min_reading = min(reading_score, na.rm = TRUE),
    max_reading = max(reading_score, na.rm = TRUE),
    
    mean_math = mean(math_score, na.rm = TRUE),
    median_math = median(math_score, na.rm = TRUE),
    sd_math = sd(math_score, na.rm = TRUE),
    min_math = min(math_score, na.rm = TRUE),
    max_math = max(math_score, na.rm = TRUE),
    
    mean_science = mean(science_score, na.rm = TRUE),
    median_science = median(science_score, na.rm = TRUE),
    sd_science = sd(science_score, na.rm = TRUE),
    min_science = min(science_score, na.rm = TRUE),
    max_science = max(science_score, na.rm = TRUE),
    
    mean_paredint = mean(paredint, na.rm = TRUE),
    median_paredint = median(paredint, na.rm = TRUE),
    sd_paredint = sd(paredint, na.rm = TRUE),
    min_paredint = min(paredint, na.rm = TRUE),
    max_paredint = max(paredint, na.rm = TRUE),
    
    mean_wealth = mean(wealth, na.rm = TRUE),
    median_wealth = median(wealth, na.rm = TRUE),
    sd_wealth = sd(wealth, na.rm = TRUE),
    min_wealth = min(wealth, na.rm = TRUE),
    max_wealth = max(wealth, na.rm = TRUE)
  ) %>%
  ungroup()

# Pivot the table to make it longer with variable as rows and statistics as columns
country_summary_stats_long <- country_summary_stats %>%
  pivot_longer(
    cols = -cnt,  # Pivot all columns except 'cnt'
    names_to = "Variable_Statistic",
    values_to = "Value"
  ) %>%
  separate(Variable_Statistic, into = c("Statistic", "Variable"), sep = "_") %>%
  pivot_wider(
    names_from = Statistic,
    values_from = Value
  )

# Print the reshaped table
print(country_summary_stats_long, n = 60)
```

### ii. Distribution of Students by Country, Country Code, OECD Membership
```{r categorical_distribution, echo=FALSE}
# Create the summary table
country_summary <- student_ten_filtered %>%
  group_by(cntryid, cnt) %>%
  summarise(Observations = n()) %>%
  ungroup()

# Print the summary table
print(country_summary)

# Print the number of unique countries
unique_countries <- nrow(country_summary)
print(paste("Number of unique countries:", unique_countries))

oecd_counts <- table(student_ten_filtered$oecd)
# Generate the message using paste
message <- paste("The OECD variable shows that there are",
                 oecd_counts["No"], "observations from non-OECD countries and",
                 oecd_counts["Yes"], "from OECD countries.")

# Print the message
print(message)

# Calculate the counts for gender
gender_counts <- table(student_ten_filtered$student_gender)

# Generate the message using paste
gender_message <- paste("The dataset includes",
                        gender_counts["Female"], "female students and",
                        gender_counts["Male"], "male students.")

# Print the message
print(gender_message)
```

### iii. Cross-tabulation of Variables
```{r cross_tabulation, echo=FALSE}
# Cross-tabulation between country and student_gender
table(student_ten_filtered$cnt, student_ten_filtered$student_gender)
```

### iv. Average Scores by Subjects for Top 6 and Bottom 4 Countries
```{r top_bottom_countries, echo=FALSE}
# Create a new column to label top and bottom countries
student_ten_filtered <- student_ten_filtered %>%
  mutate(Group = ifelse(cnt %in% c("HKG", "KOR", "TAP", "MAC", "GBR", "USA"), "Top", "Bottom"))

# Calculate the mean scores for reading, math, and science by group
group_summary <- student_ten_filtered %>%
  group_by(Group) %>%
  summarise(
    mean_reading = mean(reading_score, na.rm = TRUE),
    mean_math = mean(math_score, na.rm = TRUE),
    mean_science = mean(science_score, na.rm = TRUE)
  ) %>%
  ungroup()

# Reshape the data for plotting
group_summary_long <- group_summary %>%
  pivot_longer(
    cols = -Group,
    names_to = "Subject",
    values_to = "Mean_Score"
  ) %>%
  mutate(Subject = recode(Subject, 
                          mean_reading = "Reading", 
                          mean_math = "Math", 
                          mean_science = "Science"))

# Plot the mean scores by group
score_group <- ggplot(group_summary_long, aes(x = Subject, y = Mean_Score, fill = Group)) +
  geom_bar(stat = "identity", position = "dodge") +
  labs(title = "Mean Scores by Subject for Top and Bottom Countries",
       x = "Subject",
       y = "Mean Score") +
  theme_minimal() +
  scale_fill_manual(values = c("Top" = "blue", "Bottom" = "red"))

score_group

# Save the plot as a JPEG file
ggsave("score_group.jpeg", plot = score_group, 
       width = 10, height = 8, dpi = 600)
```

### v. Calculate Mean Scores by Group and Gender
```{r mean_scores_gender_group, echo=FALSE}
# Calculate the mean scores for reading, math, and science by group and gender
group_gender_summary <- student_ten_filtered %>%
  group_by(Group, student_gender) %>%
  summarise(
    mean_reading = mean(reading_score, na.rm = TRUE),
    mean_math = mean(math_score, na.rm = TRUE),
    mean_science = mean(science_score, na.rm = TRUE)
  ) %>%
  ungroup()
```


```{r reshape_data, echo=FALSE} 
# Reshape the data for plotting
group_gender_summary_long <- group_gender_summary %>%
  pivot_longer(
    cols = starts_with("mean_"),
    names_to = "Subject",
    values_to = "Mean_Score"
  ) %>%
  mutate(Subject = recode(Subject, 
                          mean_reading = "Reading", 
                          mean_math = "Math", 
                          mean_science = "Science"))
```


```{r plot_results, echo=FALSE}
# Plot the mean scores by group, gender, and subject
score_group_gender <- ggplot(group_gender_summary_long, aes(x = Subject, y = Mean_Score, fill = Group)) +
  geom_bar(stat = "identity", position = "dodge") +
  facet_wrap(~student_gender) +
  labs(title = "Mean Scores by Subject, Gender, and Group (Top/Bottom Countries)",
       x = "Subject",
       y = "Mean Score") +
  theme_minimal() +
  scale_fill_manual(values = c("Top" = "blue", "Bottom" = "red")) +
  theme(legend.position = "top")

score_group_gender

# Save the plot as a JPEG file
ggsave("score_group_gender.jpeg", plot = score_group_gender, 
       width = 10, height = 8, dpi = 600)
```

### vi. Calcualte Mean Scores by Group, Gender, and Wealth (Quartiles)
```{r scores_group_gender_wealth, echo=FALSE}
# Categorize wealth into quartiles
student_ten_filtered <- student_ten_filtered %>%
  group_by(Group) %>%
  mutate(wealth_quartile = ntile(wealth, 4)) %>%
  ungroup()

# Calculate the mean scores for reading, math, and science by group, gender, and wealth quartile
group_gender_wealth_summary <- student_ten_filtered %>%
  group_by(Group, student_gender, wealth_quartile) %>%
  summarise(
    mean_reading = mean(reading_score, na.rm = TRUE),
    mean_math = mean(math_score, na.rm = TRUE),
    mean_science = mean(science_score, na.rm = TRUE)
  ) %>%
  ungroup()

# Reshape the data for plotting
group_gender_wealth_summary_long <- group_gender_wealth_summary %>%
  pivot_longer(
    cols = starts_with("mean_"),
    names_to = "Subject",
    values_to = "Mean_Score"
  ) %>%
  mutate(Subject = recode(Subject, 
                          mean_reading = "Reading", 
                          mean_math = "Math", 
                          mean_science = "Science"))

# Plot the mean scores by group, gender, wealth quartile, and subject
scores_group_gender_wealth<- ggplot(group_gender_wealth_summary_long, aes(x = wealth_quartile, y = Mean_Score, fill = Group)) +
  geom_bar(stat = "identity", position = "dodge") +
  facet_grid(student_gender ~ Subject) +
  labs(title = "Mean Scores by Subject, Gender, Wealth Quartile, and Group (Top/Bottom Countries)",
       x = "Wealth Quartile",
       y = "Mean Score",
       fill = "Group") +
  theme_minimal() +
  scale_fill_manual(values = c("Top" = "blue", "Bottom" = "red")) +
  theme(legend.position = "top")

scores_group_gender_wealth

# Save the plot as a JPEG file
ggsave("scores_group_gender_wealth.jpeg", plot = scores_group_gender_wealth, 
       width = 10, height = 8, dpi = 600)
```

### vii. Scores by Group, Gender, and Parental Education
```{r scores_group_gender_paredint, echo=FALSE}
# Handle missing values by filtering them out (if necessary)
student_ten_filtered <- student_ten_filtered %>%
  filter(!is.na(paredint))

# Categorize parental education index into quartiles
student_ten_filtered <- student_ten_filtered %>%
  group_by(Group) %>%
  mutate(paredint_quartile = ntile(paredint, 4)) %>%
  ungroup()

# Calculate the mean scores for reading, math, and science by group, gender, and parental education quartile
group_gender_paredint_summary <- student_ten_filtered %>%
  group_by(Group, student_gender, paredint_quartile) %>%
  summarise(
    mean_reading = mean(reading_score, na.rm = TRUE),
    mean_math = mean(math_score, na.rm = TRUE),
    mean_science = mean(science_score, na.rm = TRUE)
  ) %>%
  ungroup()

# Reshape the data for plotting
group_gender_paredint_summary_long <- group_gender_paredint_summary %>%
  pivot_longer(
    cols = starts_with("mean_"),
    names_to = "Subject",
    values_to = "Mean_Score"
  ) %>%
  mutate(Subject = recode(Subject, 
                          mean_reading = "Reading", 
                          mean_math = "Math", 
                          mean_science = "Science"))

# Plot the mean scores by group, gender, parental education quartile, and subject
scores_group_gender_paredint <- ggplot(group_gender_paredint_summary_long, aes(x = paredint_quartile, y = Mean_Score, fill = Group)) +
  geom_bar(stat = "identity", position = "dodge") +
  facet_grid(student_gender ~ Subject) +
  labs(title = "Mean Scores by Subject, Gender, Parental Education Quartile, and Group (Top/Bottom Countries)",
       x = "Parental Education Quartile",
       y = "Mean Score",
       fill = "Group") +
  theme_minimal() +
  scale_fill_manual(values = c("Top" = "blue", "Bottom" = "red")) +
  theme(legend.position = "top")

scores_group_gender_paredint

# Save the plot as a JPEG file
ggsave("scores_group_gender_paredint.jpeg", plot = scores_group_gender_paredint, 
       width = 10, height = 8, dpi = 600)
```
